Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-06-04 paper.tex
Define the emissivity and opacity constituents earlier

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
Neutrino radiation obeys the relativistic Boltzmann Equation
\todo{point out limit of QKEs}
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f(x^\alpha;\varepsilon,\upsilon_1,\upsilon_2) = C[f],
\end{equation}
where $\frac{d}{d\lambda}$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x})
and $C[f]$ is the source term arising from interactions with the medium,
having dimension ${\rm energy}\,{\rm length}^{-1}$.
The source term varies over phase space $(x^\alpha,p_\beta)$,
and depends locally on the distribution function $f$
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For clarity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

We make the right hand side of Eqn.~\ref{eqn:boltzmann} explicit by writing
the source term linear in $f$:
\begin{equation}
  \label{eqn:boltzmann_linear}
  \frac{d}{d\lambda}f =
  \mathscr{E}^* - \mathscr{K}^* f,
\end{equation}
where we have introduced
$\mathscr{E}^*$, the invariant total emissivity, and
$\mathscr{K}^*$, the invariant total opacity,
and for simplicity we have suppressed the arguments of $f$.
These coefficients are derived by considering Fermi-blocking and
dependencies on neutrino and antineutrino distribution functions at other
momenta.
They are derived in App.~\ref{sec:source_terms}.

@@

It is customary to approximate the scattering kernel to linear order in
$\omega$ (as in~\cite[Eqn.~4.21]{shib2011-truncated_moment}):
\begin{equation}
  \label{eqn:se_kernel_expansion}
  R_{{\rm SE},i}(\varepsilon,\omega) \approx
  R_{{\rm SE},i}^0(\varepsilon) + \omega \, R_{{\rm SE},i}^1(\varepsilon),
\end{equation}
which allows us to write Eqns.~\ref{eqn:se_factor} and \ref{eqn:se_opacity} as
\todo{drop anisotropic term and defend}
\begin{align}
  \label{eqn:se_factor_expansion}
  \Phi_{{\rm SE},i}(\varepsilon) &= \frac{1}{\mathscr{K}_{{\rm SE},i}}
  \big(R_{{\rm SE},i}^0(\varepsilon) J(\varepsilon) +
   R_{{\rm SE},i}^1(\varepsilon) \ell_\alpha H^\alpha(\varepsilon) \big) \\
  \label{eqn:se_opacity_expansion}
  \mathscr{K}_{{\rm SE},i}(\varepsilon) &=
  4\pi \frac{\varepsilon^3}{(2\pi)^3} R_{{\rm SE},i}^0(\varepsilon),
\end{align}
where we have made the energy-dependence explicit.

However, we don't know the energy dependence of $J(\varepsilon)$ and
$H^\alpha(\varepsilon)$ in general,
since simulations usually employ a gray moment scheme,
evolving only the energy-integrated moments, $J$ and $H^\alpha$
and sometimes $G$.
Nevertheless, we may reconstruct these energy-dependent source terms
from the gray moments by assuming a diluted Fermi-Dirac spectrum.

We use the following procedure:
\begin{enumerate}
\item
  Interpolate the fluid temperature, $T$,
  equilibrium neutrino chemical potential, $\mu_\nu$,
  and evolved neutrino energy and number densities in the fluid frame,
  $J$ and $G$, from the simulation grid.
\item
  Compute the average neutrino energy in the fluid frame
  \begin{equation}
    \langle \varepsilon \rangle \equiv J/G.
  \end{equation}
\item
  Compute the neutrino spectral temperature, assuming equilibrium with the
  fluid
  \begin{equation}
    T_\nu = \langle \varepsilon \rangle
    \frac{\mathscr{F}_2(\mu_\nu/T)}{\mathscr{F}_3(\mu_\nu/T)},
  \end{equation}
  where $\mathscr{F}_b$ is the Fermi integral
  \begin{equation}
    \mathscr{F}_b(\eta) = \int_0^\infty dx \, x^b (1+e^{x-\eta})^{-1}.
  \end{equation}
\item
  Assume the background neutrino fields have the same total density
  as the evolved moments
  \begin{align}
    J(\varepsilon) &=
    J \frac{\varepsilon^3}{T_\nu^4 \,\mathscr{F}_3(\mu_\nu/T)}
    (1+\exp(\varepsilon/T_\nu-\mu_\nu/T)^{-1}, \\
    H^\mu(\varepsilon) &=
    H^\mu \frac{\varepsilon^3}{T_\nu^4\, \mathscr{F}_3(\mu_\nu/T)}
    (1+\exp(\varepsilon/T_\nu-\mu_\nu/T)^{-1}.
  \end{align}
\end{enumerate}

>>>>

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
Neutrino radiation obeys the relativistic Boltzmann Equation
\todo{point out limit of QKEs}
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f(x^\alpha;\varepsilon,\upsilon_1,\upsilon_2) = C[f],
\end{equation}
where $\frac{d}{d\lambda}$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x})
and $C[f]$ is the source term arising from interactions with the medium,
having dimension ${\rm energy}\,{\rm length}^{-1}$.
The source term varies over phase space $(x^\alpha,p_\beta)$,
and depends locally on the distribution function $f$
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For simplicity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

We make the right hand side of Eqn.~\ref{eqn:boltzmann} explicit by writing
the source term linear in $f$:
\begin{equation}
  \label{eqn:boltzmann_linear}
  \frac{d}{d\lambda}f =
  \mathscr{E}^* - \mathscr{K}^* f,
\end{equation}
where we have introduced
$\mathscr{E}^*$, the invariant total emissivity, and
$\mathscr{K}^*$, the invariant total opacity,
and for simplicity we have suppressed the arguments of $f$.
These coefficients are computed by considering Fermi-blocking and
dependencies on neutrino and antineutrino distribution functions at other
momenta.
We provide a detailed derivation for these terms in App.~\ref{sec:source_terms},
for the two classes of interactions we consider in this work,
absorption/emission and elastic scattering
(Tab.~\ref{tab:neutrino_processes}). Within the scope of this work, we have
\begin{align}
  \label{eqn:inv_emissivity_components}
  \mathscr{E}^* &= \mathscr{E}_{\rm AE} + \mathscr{E}_{\rm SE},\\
  \label{eqn:inv_opacity_components}
  \mathscr{K}^* &= \mathscr{K}_{\rm AE} + \mathscr{K}_{\rm SE}.
\end{align}

For absorption/emission interactions (App.~\ref{ssec:sources_ae}) these are
simply the sums over the emissivities and absorptive opacities for each
reaction $i$:
\begin{align}
  \label{eqn:ae_emissivity_summed}
  \mathscr{E}_{\rm AE}(\varepsilon)
  &= \frac{(2\pi)^3}{\varepsilon^2} \sum_{i\,{\rm reactions}} \eta_i(\varepsilon), \\
  \label{eqn:ae_opacity_summed}
  \mathscr{K}_{\rm AE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm a},i}(\varepsilon).
\end{align}

For elastic scattering interactions (App.~\ref{ssec:sources_se}) these are
the background fields and the sum over the scattering opacities for each
reaction $i$:
\begin{align}
  \label{eqn:se_emissivity_summed}
  \mathscr{E}_{\rm SE}(\varepsilon)
  &= \mathscr{K}_{\rm SE}(\varepsilon) \, \Phi_{\rm SE}(\varepsilon) \, \\
  \label{eqn:se_opacity_summed}
  \mathscr{K}_{\rm SE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm s},i}(\varepsilon).
\end{align}

This paper introduces the use of neutrino densities and fluxes evolved in a
truncated moment simulation to estimate the background field, $\Phi_{\rm SE}$.
App.~\ref{ssec:sources_se} details a method to calculate
$\Phi_{\rm SE}(\varepsilon)$ in two different ways:
\begin{itemize}
\item
  the \emph{spectral} method using densities and fluxes
  extracted from a simulation evolved over multiple energy groups
  to compute the background field with Eqn.~\ref{eqn:se_factor_expansion},
\item
  the \emph{gray} method using energy-integrated densities and fluxes
  extracted from a gray simulation and approximating the
  energy dependence with Eqns.~\ref{eqn:J_from_gray} and
  \ref{eqn:H_from_gray}.
\end{itemize}

@@

It is customary to approximate the scattering kernel to linear order in
$\omega$ (as in~\cite[Eqn.~4.21]{shib2011-truncated_moment}):
\begin{equation}
  \label{eqn:se_kernel_expansion}
  R_{{\rm SE},i}(\varepsilon,\omega) \approx
  R_{{\rm SE},i}^0(\varepsilon) + \omega \, R_{{\rm SE},i}^1(\varepsilon),
\end{equation}
which allows us to write Eqns.~\ref{eqn:se_factor} and \ref{eqn:se_opacity} as
\todo{drop anisotropic term and defend}
\begin{align}
  \label{eqn:se_factor_expansion}
  \Phi_{{\rm SE},i}(\varepsilon) &= \frac{1}{\mathscr{K}_{{\rm SE},i}}
  \big(R_{{\rm SE},i}^0(\varepsilon) J(\varepsilon) +
   R_{{\rm SE},i}^1(\varepsilon) \ell_\alpha H^\alpha(\varepsilon) \big) \\
  \label{eqn:se_opacity_expansion}
  \mathscr{K}_{{\rm SE},i}(\varepsilon) &=
  4\pi \frac{\varepsilon^3}{(2\pi)^3} R_{{\rm SE},i}^0(\varepsilon),
\end{align}
where we have made the energy-dependence explicit.
When we get $J(\varepsilon)$ and $H^\alpha(\varepsilon)$ from a multi-group
moment evolution, we call this the ``spectral'' method.

However moment evolutions with multiple energy groups are still rare.
Most simulations employ a gray moment scheme,
evolving only the energy-integrated moments, $J$ and $H^\alpha$
and sometimes the number density $G$.
We may reconstruct the energy-dependent source terms
from the gray moments by assuming a diluted Fermi-Dirac spectrum.

We use the following procedure:
\begin{enumerate}
\item
  Interpolate the fluid temperature, $T$,
  equilibrium neutrino chemical potential, $\mu_\nu$,
  and evolved neutrino energy and number densities in the fluid frame,
  $J$ and $G$, from the simulation grid.
\item
  Compute the average neutrino energy in the fluid frame
  \begin{equation}
    \langle \varepsilon \rangle \equiv J/G.
  \end{equation}
\item
  Compute the neutrino spectral temperature, assuming equilibrium with the
  fluid
  \begin{equation}
    T_\nu = \langle \varepsilon \rangle
    \frac{\mathscr{F}_2(\mu_\nu/T)}{\mathscr{F}_3(\mu_\nu/T)},
  \end{equation}
  where $\mathscr{F}_b$ is the Fermi integral
  \begin{equation}
    \mathscr{F}_b(\eta) = \int_0^\infty dx \, x^b (1+e^{x-\eta})^{-1}.
  \end{equation}
\item
  Assume the background neutrino fields have the same total density
  as the evolved moments
  \begin{align}
    \label{eqn:J_from_gray}
    J(\varepsilon) &=
    J \frac{\varepsilon^3}{T_\nu^4 \,\mathscr{F}_3(\mu_\nu/T)}
    (1+\exp(\varepsilon/T_\nu-\mu_\nu/T)^{-1}, \\
    \label{eqn:H_from_gray}
    H^\mu(\varepsilon) &=
    H^\mu \frac{\varepsilon^3}{T_\nu^4\, \mathscr{F}_3(\mu_\nu/T)}
    (1+\exp(\varepsilon/T_\nu-\mu_\nu/T)^{-1}.
  \end{align}
\end{enumerate}
When we use these moments in Eqn.~\ref{eqn:se_factor_expansion},
we call this the ``gray'' method.
